---
title: "feb04_2022 class"
author: "Rachel Diao"
date: "2/4/2022"
output:
  pdf_document: default
  html_document: default
---
# This is an R Markdown header 
## Level 2 heading 
### Level 3 heading
This is just plain text.


### We are now learning how to write functions. 

First we will write a funciton to grade some student scores. 
```{r Learning how to write functions}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

mean(student1)

#For student 2, you'd have to drop the NA or else there'd be an error
mean(student2, na.rm = TRUE)

#For student 3... you're dropping NAs so you're not accounting for how little this person did
mean(student3, na.rm = TRUE)

#Convert all NAs in a vector to 0 
student3_placehold <- student3
student3_placehold[is.na(student3)] <- 0
student3_placehold
mean(student3_placehold)
```
Now we know how to get rid of NAs, we can write a function to calculate any student's average grade. This function takes in grade_list, a character vector containing numerical scores (or NA if any assignments were omitted). 

Use if/else to approach this; if a student's grade list contains NA, then convert all the NAs to 0 before calculating the average and returning that as the output of the function. If there are no NAs, then proceed to normally calculate the mean. 
```{r Calculate average function}
calc_average <- function(grade_list){
  if (any(is.na(grade_list)) == TRUE) {
    grade_list[is.na(grade_list)] <- 0
    return(mean(grade_list))
  }
  else{
    return(mean(grade_list))
  }
}
```

Test out our function on each student
```{r Test on students}
s1_avg <- calc_average(student1)
s2_avg <- calc_average(student2)
s3_avg <- calc_average(student3)
```

Now we can try altering this function to also calculate each student's grade after they drop their lowest score. 

Function: grade 
Takes in a character vector containing numerical grades (and NAs for any omitted assignments
```{r}
#Hint: which.min() finds out which position of a list contains a minimum (same with which.max())
which.min(student1)
student1[which.min(student1)] #prints out the lowest score after finding its position

which.min(student3_placehold)
#In the case of student 3, there are multiple grades with the minimum value (0), but this function will still only pick one to proceed with. 

#Test out notation for dropping an element form a vector
student3_placehold_drop <- student3_placehold[-which.min(student3_placehold)]

grade <- function(gradelist){
  #First: if any NAs are present in the list, convert to 0
  if (any(is.na(gradelist)) == TRUE) {
    gradelist[is.na(gradelist)] <- 0
  }
  
  #Then we can find the minimum and drop it using vec_name[-element] notation
  dropped <- gradelist[-which.min(gradelist)]
  
  #Then return the average
  return(mean(dropped))
}
```

Test the 'grade' function on the students' grades. 
```{r}
grade(student1)
grade(student2)
grade(student3)
```

### Question 1
Now apply this to dataframe 'student_homework.csv'. Import into R using read.csv. Each column is for a different homework assignment and each row is for a different student. 

We want to use the function 'apply' to apply the 'grade' function to each row of this dataframe. 
```{r}
#Add argument row.names = 1 so that the first column (student names) is an index and no longer content of the dataframe.
sh <- read.csv('student_homework.csv', row.names = 1)

#Apply function to the dataframe that has no student names 
#Margin = shows how to apply the function --> i.e. 1 is rows, 2 is columns
apply(sh, MARGIN = 1, FUN = grade)
```

### Question 2 
The top-scoring student in the gradebook is student 18, with an average of 94.50.

```{r}
averages <- apply(sh, MARGIN = 1, FUN = grade)
which.max(averages)
```



### Question 3
The homework that was toughest on the students was hw2, with a class average of 72.80.
```{r}
#Duplicate the dataframe and then convert all NAs to 0
dup_sh <- sh
dup_sh[is.na(dup_sh)] <- 0

#Find the average score for each column
colMeans(dup_sh)

#Alternative method: 
apply(sh, 2, mean, na.rm=TRUE)

#BUT THE ANSWERS ARE DIFFERENT depending on whether you remove NAs or convert them to 0
#Visualize the data to figure out what's going on here
boxplot(sh)
```



