---
title: 'Project: Vaccination Rate'
author: "Rachel Diao"
date: "3/4/2022"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---

```{r}
vax <- read.csv('covid19vaccinesbyzipcode_test.csv')
#head(vax)
```

## Question 1 
Total number of people fully vaccinated is under column "persons_fully_vaccinated".

## Question 2
Zip code tabulation area is under "zip_code_tabulation_area".

## Question 3
The earliest date is 2021-01-05.

## Question 4
The latest date is 2022-03-01.

Get an overview of the dataset with skim: 
```{r}
library(skimr)
skimr::skim(vax)
```

## Question 5
There are 10 numeric columns in the dataset.

## Question 6
There are 18338 missing values in the persons_fully_vaccinated column

## Question 7
`r 18338/107604*100`% of persons_fully_vaccinated are missing.

# Working with dates

Load in package 'lubridate'!
```{r}
library(lubridate)
today()
```

Convert data in as_of_date column to lubridate format! 
```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can actually do operations on dates. To calculate difference from today - the earliest date in this dataset is 
```{r}
today() - vax$as_of_date[1]
```
Days that the dataset spans: 
```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

How many days have I been alive? - 9025 days
```{r}
today() - ymd('1997-06-18')
```

## Question 9 
Difference between today and the last date in the dataset:`r today() - vax$as_of_date[nrow(vax)]` 

## Question 10
There are 61 unique dates in the dataset (answer from the skimr summary)

# Working with zipcodes

Load in package zipcodeR! With geocode_zip(), we can get the centroid of the region any zipcode covers. 
```{r}
library(zipcodeR)
geocode_zip('92037')
```

Calculate distance (in miles) between centroids of any two zipcodes: 
```{r}
zip_distance('92037','92109')
```

Use reverse_zipcode() to pull lots of info on zipcodes:
```{r}
reverse_zipcode(c('92037', "92109") )
```


# Focus on San Diego county 
```{r}
# Subset to San Diego county only areas
sd <- vax[vax$county=='San Diego', ]
```

Can also do the same in dplyr
```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")
nrow(sd)
```

Filter for areas where population is greater than 10,000:
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```
## Question 11
There are `r length(unique(sd$zip_code_tabulation_area))` distinct zip codes in SD county. 

## Question 12 
`r sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]` has the largest 12+ population in this dataset.

# Data for 2022-03-01
```{r}
recent <- filter(vax, county == "San Diego", as_of_date=='2022-03-01')

#Average percent of population fully vaccinated in San Diego on this day
mean(recent$percent_of_population_fully_vaccinated, na.rm=TRUE)
```
## Question 13 
70.53% of the population in San Diego was vaccinated by 03-01-2022.

## Question 14
```{r}
library(ggplot2)
ggplot(recent, aes(x=percent_of_population_fully_vaccinated)) + 
  geom_histogram(bins=15) + geom_rug() +
  xlab('% Population Fully Vaccinated on 2022-03-01 (By Zip Code)') + 
  ylab('Count')
```

# UCSD data
```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```
## Question 15 
Vaccination rate time-course for UCSD zip code: 
```{r}
ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x='Date', y="% Vaccinated in UCSD/La Jolla")
```
# Compare to similarly-sized areas: 
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-03-01")
```

## Question 16
`r mean(vax.36$percent_of_population_fully_vaccinated)*100`% people are vaccinated in areas as large as 92037. 
```{r}
ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept=mean(vax.36$percent_of_population_fully_vaccinated), 
             linetype='dashed', col='red') + 
  ylim(c(0,1)) +
  labs(x='Date', y="Percent Vaccinated")
```

## Question 17 
```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

## Question 18
```{r}
ggplot(vax.36) + aes(x=percent_of_population_fully_vaccinated) + 
  geom_histogram(bins=30) + geom_rug() + 
  xlim(c(0,1)) +
  xlab('% Vaccinated by 03-01-2022') + ylab('Count')
```
## Question 19

92109 (55.20%) and 92040 (72.38%) averages are both below the average % vaccinated for all counties in California with population size similar or larger than 92037, though 92109 is only about 1% lower than the average.
```{r}
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

## Question 20 
```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color='blue') +
  ylim(c(0,1)) +
  labs(x='Date', y='% of Population Vaccinated (by Zip Code)',
       title='Vaccination Rates Across California',
       subtitle='Only areas with population above 36000 are shown') +
  geom_hline(yintercept = mean(vax.36$percent_of_population_fully_vaccinated), linetype='dashed')
```

```{r}
vax.extras <- filter(vax, zip_code_tabulation_area==c('94555', '92707', '94112'))

ggplot(vax.extras) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(color=zip_code_tabulation_area) +
  ylim(c(0,1)) +
  labs(x='Date', y='% of Population Vaccinated (by Zip Code)',
       title='Vaccination Rates Across California',
       subtitle='Only areas with population above 36000 are shown') +
  geom_hline(yintercept = mean(vax.36$percent_of_population_fully_vaccinated), linetype='dashed')
```